\thetitle

### 2021-02-23 writing/mutate-benchmarks.tex
factor out figures

The starting point for our corpus of programs
is~\citet{gtnffvf-jfp-2019}'s gradual typing benchmark suite. The
benchmark suite consists of fully typed correct programs that are written by different authors
and have been used, maintained and evolved by their authors and others over a
number of years.  They range widely in the size, complexity, purpose and
features of Typed Racket they employ.  Furthermore the benchmarks use advanced aspects
of Typed Racket's type system such as occurrence
typing~\cite{tf-icfp-2010}, types for mutable and immutable data
structures, polymorphic types, types for first-class classes and objects, and types for
Racket's numeric tower~\cite{stathff-padl-12}. Without loss of the diversity of the benchmarks,
we select the ten largest in terms of numbers of components (between 6 and 14 components).
These benchmarks also have the most complex dependency graphs and, hence,
can make the debugging process the hardest for the rational programmer.
The table in figure~\ref{table:benchmark-descriptions} shows our selection
together with a short description for each benchmark.

\begin{figure}
\begin{tabular}{p{1.75cm} | p{6cm} }
  {\bf  name} & {\bf description (author)}  \\

\hline
  \texttt{acquire} & Object-oriented board game implementation (M. Felleisen)  \\%[1em]

\hline
  \texttt{gregor} & Utilities for calendar dates (J. Zeppieri) \\%[1em]

\hline
  \texttt{kcfa} & Functional implementation of 2CFA for the lambda calculus (M. Might) \\%[1em]

\hline
  \texttt{quadT} & Converter from S-expression source code to PDF format (M. Butterick)\\%[1em]

\hline
  \texttt{quadU} & Converter from S-expression source code to PDF format  (B. Greenman) \\%[1em]

\hline
  \texttt{snake} & Functional implementation of the  Snake video game (D. Van Horn) \\%[1em]

\hline
  \texttt{suffixtree} & Algorithm for common longest subsequences between strings. (D. Yoo) \\%[1em]

\hline
  \texttt{synth} & Converter of notes and drum beats to WAV (V. St-Amour \& N. Toronto) \\%[1em]

\hline
  \texttt{take5} & Mixin-based card game simulator (M.Felleisen)  \\%[1em]

\hline
  \texttt{tetris} & Functional implementation of Tetris (D. Van Horn) \\%[1em]

\end{tabular}
  \caption{Benchmarks summary.}
  \label{table:benchmark-descriptions}
\end{figure}

>>>>

Over the past decade, \citet{gtnffvf-jfp-2019} have curated a collection of
Typed Racket programs for systematically measuring the implementation's
performance. The benchmark suite consists of fully typed, correct programs,
written by a number of different authors, who have maintained and evolved these
programs over several years. The programs range widely in size, complexity,
purpose, and in programming style. The latter is mostly a reference to their
reliance on many Typed Racket features: occurrence typing~\cite{tf-icfp-2010},
types for mutable and immutable data structures~\cite{hpst-sfp-2010}, types for
first-class classes and objects~\cite{tsdtf-oopsla-2012}, and types for Racket's
numeric tower~\cite{stathff-padl-12}.

Since mixing of typed and untyped code in Typed Racket takes place at the module
level, the experiment calls for benchmarks with a decent number of modules and a
variety of module dependency graphs. Filtering the benchmark suite using these
criteria yields ten suitable programs. Figure~\ref{table:benchmark-descriptions}
displays the selected benchmark programs and describes their basic
characteristic.

### 2021-02-23 writing/mutate-interesting.tex
factor out figures

The reader may have observed that mutators that produce 
type-level mistakes do not necessarily turn our benchmarks into
interesting debugging scenarios for the rational programmer.  The
benchmarks are fully typed, and type checking detects each mistake at
compile time.  Hence, we apply the mutators to versions, dubbed
dubbed{scenarios}, of the benchmarks without 
some  of their type annotations.  Without the annotations, the type
checker may not detect a mistake at compile time and, instead, the run
time checks
of the gradual type system signal an error.  Scenarios
with type-level mistakes that result in run time errors are the ideal
ground to exercise the rational programmer's ability to locate errors.

\begin{figure*}
  \begin{tabular}{p{2.1cm} | p{7cm}  | p{7.5cm} }
    {\bf name} & {\bf description} & {\bf example} \\
\hline

    \texttt{constant}
& Swaps a constant with another of the same value but different type.
& \texttt{5.6} $\rightarrow$ \texttt{5.6+0.0i} \\
\hline

    \texttt{deletion}
    & Deletes the result expression of a sequence.
    & \texttt{(begin x y z)} $\rightarrow$ \texttt{(begin x y)} \\
\hline

   \texttt{position}
 & Swaps two subexpressions.
  &  \texttt{(f a 42 "b" 0)} $\rightarrow$\texttt{(f a 42 0 "b")} \\
\hline

     \texttt{list}
    & Replaces \texttt{append} with \texttt{cons}.
    &  \begin{tabular}[t]{@{}l}
      \texttt{append} $\rightarrow$ \texttt{cons}
    \end{tabular}\\
\hline

    \texttt{top-level-id}
 & Swaps identifiers defined in the same module.
 &\texttt{(f x 42)} $\rightarrow$ \texttt{(g x 42)}\\
\hline

    \texttt{imported-id}
 & Swaps identifiers imported from the same module.
 &\texttt{(f x 42)} $\rightarrow$ \texttt{(g x 42)}\\
\hline

    \texttt{method-id}
 & Swaps two method identifiers.
 &\texttt{(send o f x 42)} $\rightarrow$ \texttt{(send o g x 42)}\\
\hline

    \texttt{field-id}
 & Swaps two field identifiers.
    &  \texttt{(get-field o f)} $\rightarrow$  \texttt{(get-field o g)}\\
\hline

     \texttt{class:init}
  & Swaps values of class initializers.
   &  \begin{tabular}[t]{@{}l}
     \texttt{(new c [a 5] [b "hello"])} $\rightarrow$\\
     \texttt{(new c [a "hello"] [b 5])}\\
    \end{tabular}\\
\hline

     \texttt{class:parent}
    & Replaces the parent of classes with \texttt{object\%}.
    &  \begin{tabular}[t]{@{}l}
     \texttt{(class foo\% (super-new))} $\rightarrow$\\
     \texttt{(class object\% (super-new))}\\
    \end{tabular}\\
\hline

     \texttt{class:public}
    & Makes a public method private and vice versa.
    & \begin{tabular}[t]{@{}l}
      \texttt{(class object (define/public (m x) x))} $\rightarrow$\\
      \texttt{(class object (define/private (m x) x))}\\
    \end{tabular}\\
\hline

     \texttt{class:super}
    & Removes \texttt{super-new} calls from class definitions.
    & \begin{tabular}[t]{@{}l}
      \texttt{(class foo\% (super-new))} $\rightarrow$ \\
      \texttt{(class foo\% (void))}\\
       \end{tabular}\\
\hline

     \texttt{arithmetic}
& Swaps arithmetic operators.
    & \texttt{+} $\rightarrow$ \texttt{-} \\
\hline

     \texttt{boolean}
& Swaps \texttt{and} and \texttt{or}.
& \texttt{and} $\rightarrow$ \texttt{or} \\
\hline

     \texttt{negate-cond}
    & Negates conditional test expressions.
    &  \texttt{(if (= x 0) t e)} $\rightarrow$ \texttt{(if (not (= x 0)) t e)}\\
\hline

     \texttt{force-cond}
    & Replaces conditional test expressions with \texttt{\#t}.
    & \texttt{(if (= x 0) t e)} $\rightarrow$ \texttt{(if \#t t e)}
\end{tabular}
  \caption{Summary of mutators.}
  \label{table:mutation-ops}
\end{figure*}

>>>>

Mutators that produce type-level mistakes do not necessarily turn
benchmarks into interesting debugging scenarios.

The benchmarks are fully typed, and type checking detects each mistake
at compile time.  Hence, we apply the mutators to versions, dubbed
dubbed{scenarios}, of the benchmarks without some of their type
annotations.  Without the annotations, the type checker may not detect
a mistake at compile time and, instead, the run time checks of the
gradual type system signal an error.  Scenarios with type-level
mistakes that result in run time errors are the ideal ground to
exercise the rational programmer's ability to locate errors.

### 2021-02-23 writing/mutate-mutators.tex
factor out figures

Of course, the gradual typing benchmarks have no mistakes for the rational
programmer to debug. Therefore, we follow ~\citet{lksfd-popl-2020} and we
inject bugs with mutation
analysis~\cite{jia2011analysis,demillo1978hints,lipton1971fault}.  A
mutation is a local syntactic change of the code that may
produce a bug. Usually, the outcome of mutating a program once, dubbed a
mutant, can be distinguished from the original program with a test case.
The better the test suite of a program, the more mutants it can ``kill.''
Mutants that are hard to ``kill'' usually correspond to subtle programming
errors, and mutation frameworks typically come with standard mutators
carefully designed to create such mutants.

Unlike the hard-to-kill mistakes introduced by standard mutators, 
gradual type systems catch shallow errors.  Thus, in order to
exercise the rational programmer in the context of gradual typing, 
we need to develop alternative mutators
that are fine-tuned to produce the type-level mistakes gradual type systems 
can detect. Our inspiration is our
more-than-a-decade long experience of making type-level mistakes in Typed
Racket. Some of these mistakes often take non-trivial effort to debug and
have become the seed for our mutators, which we summarize in
figure~\ref{table:mutation-ops}.  For each one, the figure provides a
short description and an example mutation.

Most of our mutators are self-explanatory.
The first twelve apply to most gradually typed languages, including those with classes.
The last four mutators target distinguishing features of Typed Racket.
Specifically, \texttt{arithmetic} may replace a \texttt{+} with a \texttt{-}
in an attempt to change the type of the result of the arithmetic
operation. In Typed Racket, \texttt{+}'s result is a
\texttt{Positive-Integer} when all arguments are
positive integers. However, the result of \texttt{-} is
\texttt{Integer}. In the same spirit, \texttt{boolean} aims to take
advantage of Typed Racket's ``truthiness.'' Finally, \texttt{negate-cond} and \texttt{force-cond}
attempt to confuse occurrence typing.

>>>>

Generating debugging scenarios from the benchmarks demands the design
of novel mutators. Generally speaking, a
mutator~\cite{jia2011analysis,demillo1978hints,lipton1971fault}
performs a localized syntactic change to a code basis. For the
evaluation of a blame strategy, the mutator must produce type-level
mistakes that the run-time checks of gradual typing systems or the
safety checks of the underlying language can detect. Once detected,
the rational programmer can attempt to detect the mistake by gradually
adding types to blamed modules.

Figure~\ref{table:mutation-ops} describes the sixteen mutators used
for the data reported here. They are inspired by the authors
decades-long experience of making type-level mistakes in Typed Racket,
some of which take a non-trivial effort to debug.  For each mutaror,
the figure provides a short description and an example mutation.

Most of these mutators are self-explanatory.  The first twelve apply
to most gradually typed languages, including those with classes.  The
last four target distinguishing features of Typed Racket.
Specifically, {\tt arithmetic} may replace a {\tt +} with a {\tt -} in
an attempt to change the type of the result of the arithmetic
operation. In Typed Racket, {\tt +}'s result is a {\tt
Positive-Integer} when all arguments are positive
integers~\cite{stathff-padl-12}. However, the result of {\tt -}
is {\tt Integer}. In the same spirit, the {\tt boolean} mutator aims
to take advantage of Typed Racket faithfulness to Racket's
``falsiness.'' Finally, {\tt negate-cond} and {\tt force-cond} attempt
to confuse occurrence typing.

### 2021-02-23 writing/mutate.tex
factor mutate
factor out figures

As section~\ref{sec:challenges} explains, the first challenge is to
generate a representative mistakes in representative software. The
following two subsections explain in reverse order.

@@

\sub{benchmarks} {Which Software to Use As Benchmarks}
\sub{interesting}{Are These Mutators Interesting?}

>>>>

As section~\ref{sec:challenges} explains, the first challenge is to
generate a representative mistakes in representative software. This
section explains the specific choices made for the first application
of the method: the source and nature of the chosen benchmarks
(sec.~\ref{sub:mutate-benchmarks}), the newly developed mutators
(sec.~\ref{sub:mutate-mutators}), and why they are suitable for this
experiment (sec.~\ref{sub:mutate-interesting}).

@@

\input{mutate-benchmarks-figure.tex}
\sub{benchmarks} {Which Software to Use As Benchmarks}
\sub{mutators}   {How to Mutate the Benchmarks} 
\input{mutate-mutators-figure.tex}
\sub{interesting}{Are These Mutators Interesting}
